The Evolution of Phenotypically Invariant Genetic Networks

### 2017-07-31 handnotes.tex
metabolic example with Kalman decomp

  \section*{Additional Examples}

  \subsection*{Metabolic network}

@@

  Minimally such a system may have the architecture,
  \begin{align*}
    \mathcal{S} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
      \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
  \end{align*}
  $\mathcal{S}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.

@@

  We can see that changing coordinates on this system (with any $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.

@@

     So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $A_{\min}$:
     \begin{align*}
       A' &= \begin{bmatrix}
         1.6923  &  1.5385 &  -2.6154 &  -2 \\
         0.8462  &  0.7692 &  -1.3077 &  -1 \\
         1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
         0.4231  &  0.3846 &  -0.6538 &  -0.5
       \end{bmatrix} \\
       B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
         C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
         -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
     \end{align*}

     Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, any $n$-dimensional system can be constructed using this method -- applying a change of coordinates to the Kalman decomposition -- to construct a mechanistically different system with identical phenotypic dynamics. Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architecutres to be biologically appropriate. 
%    \subsection*{Metabolic Network}
%  \begin{align*}
%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
%    \end{align*}
%    \begin{align*}
%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
%    \end{align*}
%
%    \begin{align*}
%      CV &= C \\
%      VB &= B \\
%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
%    \end{align*}
%    The GAL regulatory network in \emph{S. cerevisiae} is modelled here. The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose. The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes. This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
%
%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1. $B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations. $C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
%
%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant. These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable. Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
%
%    \subsection*{Gap Gene Network}

>>>>

  \begin{example}[Metabolic network]

@@

  Minimally such a system may have the architecture,
  \begin{align*}
    \mathcal{S}_{\min} = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 0 & -1 \\ 0 & 0 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) + \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} u_{1} \\ u_{2} \end{bmatrix} (t) \\[11pt]
      \phi(t) &= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix} \begin{bmatrix} \kappa_{1} \\ \kappa_{2} \end{bmatrix} (t) \end{array} \right.
  \end{align*}
    $\mathcal{S}_{\min}$ is minimal as its reachability and observability matrices both have rank $=2$, $\text{rank}(\mathcal{R}) = \text{rank}(\mathcal{O}) = 2 = n \implies \min$.

@@

  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.

@@

     So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $\mathcal{S}_{\min}$:
     \begin{align*}
       A' &= \begin{bmatrix}
         1.6923  &  1.5385 &  -2.6154 &  -2 \\
         0.8462  &  0.7692 &  -1.3077 &  -1 \\
         1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
         0.4231  &  0.3846 &  -0.6538 &  -0.5
       \end{bmatrix} \\
       B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
         C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
         -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
     \end{align*}

     Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, any $n$-dimensional system can be constructed using this method -- applying a change of coordinates to the Kalman decomposition -- to construct a mechanistically different system with identical phenotypic dynamics. Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architecutres to be biologically appropriate. 
  \end{example}
%    \subsection*{Metabolic Network}
%  \begin{align*}
%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
%    \end{align*}
%    \begin{align*}
%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
%    \end{align*}
%
%    \begin{align*}
%      CV &= C \\
%      VB &= B \\
%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
%    \end{align*}
%    The GAL regulatory network in \emph{S. cerevisiae} is modelled here. The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose. The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes. This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
%
%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1. $B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations. $C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
%
%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant. These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable. Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
%
%    \subsection*{Gap Gene Network}
